Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-03 MIMOSA.tex
cut cut cut

 \author{GREG FINAK$^{1\ast}$, 
 ANDREW MCDAVID$^{1}$,
 PRATIP CHATTOPADHYAY$^{3}$,
 MARIA DOMINQUEZ$^{3}$,
 STEVE DE ROSA$^{1,2}$,
 MARIO ROEDERER$^{3}$,\\
 RAPHAEL GOTTARDO$^{1}$
 \\[4pt]
\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 \\[2pt]
 {gfinak@fhcrc.org}}

@@

\begin{abstract}
{Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after stimulation), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when accessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulations, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
\end{abstract}

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous. Individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures (\textit{e.g.,} gene expression of bulk samples). For this reason, single-cell assays (\textit{e.g.,} flow, mass cytometry, single-cell gene expression) are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple (from tens to 100's) of biomarkers (\textit{e.g.,} DNA content, RNA or protein expression levels, or protein phosphorylation) simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of hematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}. \citet{McDavid2012} review the special considerations for analyzing single cell assay data. In general, special treatment of single-cell level data is warranted in order to preserve information about cell population heterogeneity.
%
%Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
%This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
%For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
%In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
%Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
%More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. DNA content, cell-surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
%Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
%A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
%While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
%This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
%Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
%For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
%These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During data analysis, cell-level biomarker measurements are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations.
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
This is the case for the Fluidigm qPCR-based single-cell gene expression technology, where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level~\citep{McDavid2012}.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$, where $N$ is the number of cells recorded and $K$ is number of biomarkers.
Using this matrix, one can form $2^K$ putative cell subsets obtained as Boolean combinations.
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
A common statistical problem is, for a given biomarker combination, to identify subjects for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).

%Another motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
A motivating example from vaccine research is the need to assess the immunogenicity of a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
 While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
 After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}. Measuring these antigen--specific T-cell subpopulations is a challenge because they
 constitute a very small fraction of the total number of CD4 and CD8 T-cells (on the order of 0.01\%--0.1\% for HIV vaccine studies like the one considered here).

 The intracellular cytokine staining (ICS) assay can measure the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that the rare cell populations and small biological effect sizes can be reliably detected.
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted.

These counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. Subjects who generate a response after stimulation are called \textit{responders}, whereas subjects that do not show any differences are called \textit{non-responders}. 
Vaccine--specific response is determined when a subject has no response at a pre--vaccine time point and a response at a post--vaccine time point. 
Specificity is further assessed by measuring response in placebo recipients. Importantly, to avoid confounding vaccine effect with time of blood draw and because baseline samples are not always available, comparisons between stimulated and unstimulated samples are typically done within a time point rather than directly between pre--and post--vaccine time points. 
%In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.

@@

In the remainder of the paper, we use the following notation to describe our model.  We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a biomarker. Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations. We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples, respectively. We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample, respectively. For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

We consider two types of immunological single-cell assays: flow cytometry and single-cell gene expression, as described below.

\textit{Flow cytometry}:
The primary dataset used here is an ICS data set generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen (prime at zero and two months, boost at four and six months)~\citep{Goepfert:2011ci}.
The goal of this data set was to assess the immune response to the vaccine across multiple antigen stimulations, time points, cytokines and T-cell subsets.
Here, we analyze a subset of the data consisting of 98 subjects from the vaccine group at two time points: day 0 and day 182.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured for each subject and time point, each with and without antigen stimulation (here we focus on the HIV Envelope peptide pool). For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken before vaccination and no response is expected there. Conversely, day 182 (26 weeks) is close to the immunogenicity peak, and many subjects are expected to respond, for some cytokines at least. In this data set a median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively). The effect sizes (differences in the proportions of stimulated and unstimulated cells) were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on the cytokine, with a few samples exhibiting larger numbers of positive events.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from stimulated samples  was compared to paired, unstimulated controls. In this data set, we have approximately 90 single cells per subject per stimulation condition with 96 gene expression measurements per cell.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like the ones presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

@@

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the un-stimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.

\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and un-stimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models,
${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p^{(s)}>p^{(u)}$. This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions, as follows:
 \begin{align*}
(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
 \end{align*}
where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are explicitly shared across the two models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using an Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
As an alternative, we also describe a fully Bayesian model, where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.

>>>>

 \author{GREG FINAK$^{1\ast}$, 
 ANDREW MCDAVID$^{1}$,
 PRATIP CHATTOPADHYAY$^{3}$,
 MARIA DOMINGUEZ$^{3}$,
 STEVE DE ROSA$^{1,2}$,
 MARIO ROEDERER$^{3}$,\\
 RAPHAEL GOTTARDO$^{1}$
 \\[4pt]
\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 \\[2pt]
 {gfinak@fhcrc.org}}

@@

\begin{abstract}
{Blood and tissue are composed of many functionally distinct cell subsets. In immunological studies, these can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on assays that provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify biomarkers (or combinations of biomarkers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after stimulation), where expression is defined as the proportion of cells expressing that biomarker (or biomarker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential biomarker expression using single-cell assays. Our model allows the inference to be subject specific, as is typically required when assessing vaccine responses, while borrowing strength across subjects through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against classical approaches for single-cell assays including Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulations, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we demonstrate how our approach can be extended to testing multivariate differential expression across multiple biomarker combinations using a Dirichlet-multinomial model and illustrate this approach using single-cell gene expression data and simulations.}{Expectations-Maximization, Markov Chain Monte Carlo,
Marginal Likelihood, Bayesian Modeling, Hierarchical Modeling,
Immunology, Flow Cytometry, Single-Cell Gene Expression, MIMOSA}
\end{abstract}

\clearpage
\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous. Individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures (\textit{e.g.,} bulk gene expression). For this reason, single-cell assays (\textit{e.g.,} flow, mass cytometry, single-cell gene expression) are an important tool in immunology, providing a functional snapshot of the immune system at a given time. These assays typically measure multiple biomarkers (\textit{e.g.,} DNA content, RNA or protein expression levels, or protein phosphorylation) simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of hematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}. \citet{McDavid2012} review the special considerations for analyzing single cell assay data. In general, special treatment of single-cell level data is warranted in order to preserve information about cell population heterogeneity.
%
%Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them.
%This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures.
%For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins.
%In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM).
%Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular biomarkers (up to 17 biomarkers).
%More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters (i.e. DNA content, cell-surface or intracellular protein levels, protein phosphorylation, RNA--level gene expression, etc.) per cell~\citep{Bendall:2011wf}.
%Although cells sorted using well-established surface biomarkers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells.
%A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}.
%While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}.
%This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data.
%Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general.
%For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time.
%These assays typically measure multiple biomarkers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood mononuclear cells  (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

Typically, cell-level measurements are thresholded as positive or negative so that subsets with different multivariate $+/-$ biomarkers can be obtained as Boolean combinations.
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others, thresholds are given by the assay technology.
This is the case for the Fluidigm's qPCR-based single-cell gene expression technology, where genes are recorded as expressed or not, at the single-cell level~\citep{McDavid2012}.
 After thresholding, we obtain a Boolean matrix of $N$ cells $\times$ $K$ biomarkers and we can form $2^K$ putative cell subsets.
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty.
A common statistical problem is to identify subjects for whom the proportion of cells expressing a specific combination is significantly different between two experimental conditions (\textit{e.g.}, before and after stimulation).

%Another motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
A motivating example from vaccine research is the need to assess the immunogenicity of a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
T-cells that recognize the antigen become \emph{activated}, produce cytokines that potentiate the immune response, and proliferate and persist in the immune system providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}. This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring cytokine production in antigen-specific T-cells in response to antigen stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 %While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
 %After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.

The antigen--specific subpopulations constitute a very small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%).  
% The intracellular cytokine staining (ICS) assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring intracellular cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
%Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
Thus, a sufficiently large number of cells must be collected (on the order of 50,000 to 100,000 T--cells) to ensure that changes in these rare cell populations can be reliably detected.
Subsequently, each individual cell is classified as either positive or negative for each cytokine based on predetermined thresholds, and the number of cells matching each subpopulation phenotype is counted.

Cell counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. \emph{Responders} are subjects whose T-cells exhibit a response to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
%Response is vaccine--specific when a subject responds at a post--vaccine time point but not at a pre--vaccine time point. 
%Response in placebo recipients is further used to assess vaccine-specificity. 
These comparisons are typically performed within a time point  in order to avoid confounding vaccine effects with time of blood draw and because baseline samples are not always available. 
%In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.

@@

We use the following notation to describe our model.  We assume that we observe cell counts from $I$ subjects in two conditions: stimulated and unstimulated. Each cell can either be positive or negative for a biomarker. Given a set of $K$ biomarkers, the measured cells can be classified into $2^K$ positive/negative biomarker combinations. We denote by $n^{(s)}_{ik}$ and $n^{(u)}_{ik}$, $i=1,\dots,I,k=1,\dots, 2^K$, the observed counts for the $2^K$ combinations in the stimulated and unstimulated samples, respectively. We denote by $N^{(s)}_{i}=\sum_k n^{(s)}_{ik}$ and $N^{(u)}_{i}=\sum_k n^{(u)}_{ik}$ the total number of cells measured for subject $i$ in each sample. For ease of notation, we denote by $\mathbf{y}_i$ the vector of observed counts for subject $i$, \textit{i.e.}, $\mathbf{y}_{i}=(\mathbf{n}^{(s)}_{i}, \mathbf{n}^{(u)}_{i})$ where $\mathbf{n}^{(s)}_{i}=\{n^{(s)}_{ik}: k=1,\dots,2^K\}$ and $\mathbf{n}^{(u)}_{i}=\{n^{(u)}_{ik}: k=1,\dots,2^K\}$, and finally $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

We consider two types of immunological single-cell assays.

\textit{Flow cytometry}:
Our primary dataset is ICS data generated as part of a  trial testing the GeoVax DNA and MVA (Modified Vaccinia Ankara) HIV vaccine in a prime-boost regimen~\citep{Goepfert:2011ci}.
The study goal was to assess the immunogenicity of different prime-boost regimens.
Here, we analyze a subset of the data consisting of 98 vaccine and placebo recipients at days 0 (pre-vaccine) and 182 (two weeks post-vaccine), focusing on CD4+ T-cell response to stimulation with HIV Envelope peptide.
Three cytokines (IFN-$\gamma$, TNF$\alpha$ and IL2) were measured for each subject and time point, each with and without antigen stimulation. No response is expected at day 0, while day 182 is close to the immunogenicity peak and many vaccinees are expected to respond for at least some cytokines. A median of 51K and 58K T--cells were collected for the stimulated and unstimulated samples, respectively (IQR $\approx$ 37K and 43K, respectively). The empirical differences in the proportions of stimulated and unstimulated cells were on the order of $10^{-4}$, with the number of positive cells typically ranging from 1--70, depending on cytokine subset.

\textit{Fluidigm single-cell gene expression}: Tetramer-sorted CD8$+$ T-cells from sixteen subjects. T-cells isolated by flow cytometry from sixteen subjects were stimulated in blocks of four subjects with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, and CMV pp65 nlv5) and gene expression post-stimulation was measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from stimulated samples  was compared to paired, unstimulated controls. In this data set, we have approximately 90 single cells per subject per stimulation condition with 96 gene expression measurements per cell.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like those presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the unstimulated and stimulated samples (Table \ref{tab:twobytwo}).

@@

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the unstimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.

\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models,
${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}\quad \text{and}\quad {\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and unstimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p^{(s)}>p^{(u)}$. This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.

\subsection{Priors}
\label{ss:priors}
Our model shares information across all subjects using exchangeable Beta priors on the unknown proportions. For non-responders, $(p^{(u)}_{i} | z_{i}=0)  \sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})$, and for responders $(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  \sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)})$, %, as follows:
% \begin{align*}
%(p^{(u)}_{i} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha^{(u)}, \beta^{(u)})\\
%(p^{(u)}_{i}|z_{i}=1) \sim \mathrm{Beta}(\alpha^{(u)},
%\beta^{(u)}) \quad\mathrm{and}\quad (p^{(s)}_{i} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha^{(s)},\beta^{(s)}),
 %\end{align*}
 where $z_i$ is an indicator variable equal to one if subject $i$ is a responder, \textit{i.e.}, ${\cal M}_1$ is true, and zero otherwise, and $\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}$ are unknown hyper-parameters shared across all subjects.
Note that the parameters $\alpha^{(u)} \text{ and } \beta^{(u)}$ are  shared across both models, whereas $\alpha^{(s)} \text{ and } \beta^{(s)}$ are only present in the alternative model.
Finally, we assume that $z_i\sim \mathrm{Be}(w)$ are independent draws from a Bernoulli distribution with probability $w$, where $w$ represents the (unknown) proportion of responders.
It follows that marginally, \textit{i.e.}, after integrating $z_i$, the $p^{(u)}_{i}$ and $p^{(s)}_{i}$ are then jointly distributed as a mixture of a one dimensional Beta distribution and a product of two Beta distributions (with a possible constraint), with mixing parameter $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha^{(u)}, \beta^{(u)}, \alpha^{(s)},\beta^{(s)}, w)$ can be estimated in an Empirical-Bayes fashion using an Expectation-Maximization algorithm~\citep{Dempster:1977ul} as described in Section \ref{s:estimation}.
We also describe a fully Bayesian model where the hyperparameters $\alpha^{(u)}, \beta^{(u)}$, $\alpha^{(s)},\text{ and }\beta^{(s)}$ are each given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1.
In this case, all parameters will be estimated via a Markov chain Monte Carlo algorithm as described in Section \ref{s:estimation}.
